---
title: "NIMBLE Example"
output: html_notebook
---

This notebook presents an abbreviated reworking of the Pump Failure model presented in Chapter 2 of the [NIMBLE Manual](https://r-nimble.org/manuals/NimbleUserManual.pdf). Note that although [NIMBLE](https://r-nimble.org/) is an R package, listed on CRAN, that installs like any other R package, you must have a C++ compiler and the standard `make` utility installed on your system before installing NIMBLE. See Chapter 4 of the manual for platform-specific installation instructions

### Pump Failure Model
The Pump Failure model is dsicussed by George et al. 
in their 1993 paper: [Conjugate Likelihood Distributions](https://www.jstor.org/stable/4616270?seq=1#page_scan_tab_contents). The paper examines Bayesian models that use conjugate priors, but which do not have closed form solutions when prior distributions are associated with the hyperparameters. The [BUGS](http://www.openbugs.net/w/FrontPage) solution of the problem is given [here](http://www.openbugs.net/Examples/Pumps.html).


The data driving the model are: `x[i]` the number of failures for pump `i` in a time interval `t[i]` where `i` goes from 1 to 10.

```{r}
library(nimble)
library(igraph)
library(tidyverse)

pumpConsts <- list(N = 10,
t = c(94.3, 15.7, 62.9, 126, 5.24,31.4, 1.05, 1.05, 2.1, 10.5))

pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))

```

Arrival times as a Poisson distribution with parameter `lambda` where `lambda` is itself modeled as a Gamma distribution with hyperparameters `alpha` and `beta`.

The model is expressed in the `BUGS` language wrapped inside the NIMBLE function `nimbleCode()` which turns the `BUGS` code into a object that can be operated on by `nimbleModel()`

```{r}
pumpCode <- nimbleCode(
  {
for (i in 1:N){
  theta[i] ~ dgamma(alpha,beta)
  lambda[i] <- theta[i]*t[i]
  x[i] ~ dpois(lambda[i])
  }
alpha ~ dexp(1.0)
beta ~ dgamma(0.1,1.0)
})

pumpInits <- list(alpha = 1, beta = 1,
theta = rep(0.1, pumpConsts$N))

```

`nimbleModel()` produces the model object that can be executed by `R` or compiled.

```{r}
pump <- nimbleModel(code = pumpCode, name = "pump", constants = pumpConsts,
                    data = pumpData, inits = pumpInits)
```

The following command lets us look at the nodes that comprise the model's directed graph and plot it.

```{r}
pump$getNodeNames()
```

```{r}
pump$plotGraph()
```


We can look at the values stored at each node. The node for `x` contains the initial values we entered into the model and the nodes for `theta` and `lambda` contain the initial calculated values

```{r}
pump$x
```


```{r}
pump$theta
```

```{r}
pump$lambda
```

We can also look at the log probabilities of the likelihoods.

```{r}
pump$logProb_x
```

Next, we use the model to simulate new values for `theta` and update the variables.

```{r}
set.seed(1)
pump$simulate("theta")
pump$theta
```

These new values will, of course, lead to new values of `lambda` and the log probabilities.

```{r}
pump$lambda
```

```{r}
pump$logProb_x
```

We can also compile the model. The C++ code generated is loaded back into R with an object that can be examined like the uncompiled model. 

```{r}
Cpump <- compileNimble(pump)
Cpump$theta
```


Now, we use the default NIMBLE MCMC algorithm to generate posterior samples from the distributions for the model parameters `alpha`, `beta`, `theta` and `lambda` along with summary statistics, and the value of Wantanabi's [WAIC](https://www.rdocumentation.org/packages/LaplacesDemon/versions/16.1.0/topics/WAIC) statistic. 

```{r}
mcmc.out <- nimbleMCMC(code = pumpCode, constants = pumpConsts,
                       data = pumpData, inits = pumpInits,
                       monitors=c("alpha","beta","theta","lambda"),
                       nchains = 2, niter = 10000,
                       summary = TRUE, WAIC = TRUE)
names(mcmc.out)
```


```{r}
mcmc.out$summary
```

```{r}
mcmc.out$WAI

```

Here, we select sample values for the parameters for pump 1 in the first chain and put them into a data frame for plotting.
```{r}
df <- data.frame(mcmc.out$samples$chain1)
df_l <- df %>% select(alpha, beta, theta.1., lambda.1.) %>% gather(key="parameter", value="value")
``` 

We plot the sample values.
```{r}
ps <- df_l %>% ggplot(aes(x=seq_along(value), y = value)) + geom_line()
ps + facet_wrap(~parameter, scales = "free")

```

And, we plot histograms.
```{r}
p <- ggplot(df_l,aes(value)) + geom_histogram(aes( y= ..density..),bins = 60) 
p + facet_wrap(~parameter, scales = "free")
               
```


Note that it is also possible to perform the MCMC simulation using the compiled model.


```{r}
mcmc.out_c<- nimbleMCMC(model=Cpump,
                       monitors=c("alpha","beta","theta","lambda"),
                       nchains = 2, niter = 10000,
                       summary = TRUE, WAIC = TRUE)
```

```{r}
mcmc.out_c$summary
```

### Monte Carlo Expectation Analysis   

Finally, to illustrate that NIMBLE is more than just an MCMC engine we NIMBLE's built-in Monte Carlo Expectation algorithm to maxime the marginal likelihood for `alpha` and `beta`. First, we create a new model and set up the "box constraints". Then, the `buildMCEM()` function is used to construct an MCEM algorithm from a NIMBLE model.

```{r}
pump2 <- pump$newModel()
box = list( list(c("alpha","beta"), c(0, Inf)))
pumpMCEM <- buildMCEM(model = pump2, latentNodes = "theta[1:10]", boxConstraints = box)
```  

Now, we run the model.

```{r}
pumpMLE <- pumpMCEM$run()
```

The authors of the NIMBLE manual point out that these results are within 0.01 of the values obtained by Geroge et al..



